---
title: 'HW2: PCA'
author: "John Han"
date: "2026-09-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## My Compression Code

```{r}
library(readr)

compress = function(X, k) {

  pca = prcomp(X, center = FALSE, scale. = FALSE)
  
  U = pca$rotation
  Z = pca$x
  
  X_k = Z[, 1:k, drop = FALSE] %*% t(U[, 1:k, drop = FALSE])
  
  error_k = norm(X - X_k, type = "F")
  
  list(approx = X_k, error = error_k, first_eig = U[,1], first_pcs = Z[,1])
}
```

## Image 1

### Read in data

```{r message=FALSE}
x <- read_csv("data/image1.csv")
x <- data.matrix(x)
```

### Original and first 10 k's

```{r}
ks = 1:10
all_approx = vector("list", 10)
for(i in ks){
  comp = compress(x, i)
  all_approx[[i]] = comp$approx
}

par(mfrow = c(4, 3), mar = c(1,1,2,1))
image(x, main = "Original", axes = F)
for(i in ks){
  image(all_approx[[i]], main = paste("k = ", i), axes = F)
}

```

### Error vs. k's

```{r}
p = dim(x)[2]
all_k = 1:p
all_error = numeric(p)
for(i in all_k){
  comp = compress(x, i)
  all_error[i] = comp$error
}

plot(all_k, all_error)

```


### First Eigenvalues and PC Scores

```{r}
first = compress(x, 1)
plot(first$first_eig)
plot(first$first_pcs)

```

### Which k is "sufficient"?

Since the first eigenvalue chart shows only one index that is not close to zero, we can argue that even k=1 is sufficient enough approximation. To further support this k value, we can see that the error vs. k graph shows a linear relationship until the very end of the k values. This means that an increase in k-value reduces the error roughly the same amount every step. So I believe even at k=1, the approximation shows a sufficiently good approximation of the original image. 

## Image 2

### Read in data

```{r message=FALSE}
x <- read_csv("data/image2.csv")
x <- data.matrix(x)
```

### Original and first 10 k's

```{r}
ks = 1:10
all_approx = vector("list", 10)
for(i in ks){
  comp = compress(x, i)
  all_approx[[i]] = comp$approx
}

par(mfrow = c(4, 3), mar = c(1,1,2,1))
image(x, main = "Original", axes = F)
for(i in ks){
  image(all_approx[[i]], main = paste("k = ", i), axes = F)
}

```

### Error vs. k's

```{r}
p = dim(x)[2]
all_k = 1:p
all_error = numeric(p)
for(i in all_k){
  comp = compress(x, i)
  all_error[i] = comp$error
}

plot(all_k, all_error)

```


### First Eigenvalues and PC Scores

```{r}
first = compress(x, 1)
plot(first$first_eig)
plot(first$first_pcs)

```

### Which k is "sufficient"?

Since the first eigenvalue chart shows only one index that is not close to zero, we can argue that even k=1 is sufficient enough approximation. To further support this k value, we can see that the error vs. k graph shows a linear relationship between k value and the error. This means that an increase in k-value reduces the error roughly the same amount every step. So I believe even at k=1, the approximation shows a sufficiently good approximation of the original image. 

## Image 3

### Read in data

```{r message=FALSE}
x <- read_csv("data/image3.csv")
x <- data.matrix(x)
```

### Original and first 10 k's

```{r}
ks = 1:10
all_approx = vector("list", 10)
for(i in ks){
  comp = compress(x, i)
  all_approx[[i]] = comp$approx
}

par(mfrow = c(4, 3), mar = c(1,1,2,1))
image(x, main = "Original", axes = F)
for(i in ks){
  image(all_approx[[i]], main = paste("k = ", i), axes = F)
}

```

### Error vs. k's

```{r}
p = dim(x)[2]
all_k = 1:p
all_error = numeric(p)
for(i in all_k){
  comp = compress(x, i)
  all_error[i] = comp$error
}

plot(all_k, all_error)
```


### First Eigenvalues and PC Scores

```{r}
first = compress(x, 1)
plot(first$first_eig)
plot(first$first_pcs)

```

### Which k is "sufficient"?

Looking at the error vs. k graph we can see an elbow of the curve that tells us that anything beyond the elbow decreases in the amount the error is reduced. The elbow is typically a nice fair point for a sufficiently good approximation. In this case, I think a k=20 is a sufficiently good approximation of the original image.


## Image 4

### Read in data

```{r message=FALSE}
x <- read_csv("data/image4.csv")
x <- data.matrix(x)
```

### Original and first 10 k's

```{r}
ks = 1:10
all_approx = vector("list", 10)
for(i in ks){
  comp = compress(x, i)
  all_approx[[i]] = comp$approx
}

par(mfrow = c(4, 3), mar = c(1,1,2,1))
image(x, main = "Original", axes = F)
for(i in ks){
  image(all_approx[[i]], main = paste("k = ", i), axes = F)
}

```

### Error vs. k's

```{r}
p = dim(x)[2]
all_k = 1:p
all_error = numeric(p)
for(i in all_k){
  comp = compress(x, i)
  all_error[i] = comp$error
}

plot(all_k, all_error)

```


### First Eigenvalues and PC Scores

```{r}
first = compress(x, 1)
plot(first$first_eig)
plot(first$first_pcs)

```

### Which k is "sufficient"?

Looking at the error vs. k graph we can see an elbow of the curve that tells us that anything beyond the elbow decreases in the amount the error is reduced. The elbow is typically a nice fair point for a sufficiently good approximation. In this case, I think a k=12 is a sufficiently good approximation of the original image.